---
title: "Overview of the Capstone Earthquake Visualization package"
author: "Daniel Languedoc"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, echo = FALSE, include = FALSE}
library(Capstone)
library(dplyr)
library(ggplot2)
library(readr)
```


The Capstone package provides the user with visualization tools that use `ggplot2` and `leaflet` to analyze NOAA earthquake data (https://www.ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1) 

## Package Functions

The Capstone package exposes the follwoing functions to users:

- `eq_clean_data()`
- `geom_timeline()`
- `geom_timeline_label()`
- `theme_timeline()`
- `eq_create_label()`
- `eq_map()`

## Tidying the data

The NOAA raw data needs to be cleaned to be efficiently processed.  `eq_clean_data` does this in the following manner

It first replaces NA values in month and day with 1 to give a complete date.
It then mutates the column like this
- creates a new Date class column called "DATE" by combining the YEAR, MONTH and DAY field
- Converts the LATTITUDE and LONGITUDE columns to numeric
- Strips the country name from the location column by calling `eq_location_clean`

The following example creates a data frame with both `eq_clean_data` and `eq_clean_location` applied

```{r eq_read_example, message = FALSE}
filename <- system.file("extdata","signif.txt", package = "Capstone")
data <- readr::read_delim(filename, delim = "\t")

eq_clean_data(data)
```


## Earthquake timelines

Capstone defines two new geoms to use with the `ggplot2` package to visualise earthquake timelines.  

`geom_timeline()` takes a clean dataframe from `eq_clean_data` and returns a basic time of when the eathequakes occurred.  The size of the plot markers is related to the magnitude of the earthquake and the color to the number of reported deaths. The "x" aesthetic is required and maps to the x-axis of the timeline.  The "y" easthetic is optional and can be used to group data by a variable, i.e. country

`geom_timeline_label()` adds text labels to the time line.  The labels are tilted to 45 degrees.  The "label" aesthic is required and should contain the name of a column that contains the information to display.

As well as these 2 primary geoms, the package includes `theme_timeline` which tidies up the display by moving the legend to the right to avoid overlapping labels

```{r eq_timeline_example, message = FALSE}
filename <- system.file("extdata","signif.txt", package = "Capstone")
data <- readr::read_delim(filename, delim = "\t")

eq_clean<-eq_clean_data(data)

eq_clean %>% filter(COUNTRY %in% c("CANADA", "MEXICO"), YEAR > 2000) %>%
  ggplot(aes(x = DATE,
             y = COUNTRY,
             color = as.numeric(TOTAL_DEATHS),
             size = as.numeric(EQ_PRIMARY)
  )) +
  geom_timeline() +
  geom_timeline_label(aes(label = LOCATION_NAME), maxCountries = 5) +
  theme_timeline() +
  labs(title = "Earthquake Fatalities", color = "# deaths", size = "Richter scale value") +
  scale_color_gradient(low = "black", high = "red")
```

## EarthQuake Maps

There are also functions to display geographical data using the `leaflet` package.

`eq_map()` takes a clean dataframe from `eq_clean_data` and returns a map of the affected area.  The map is automatically trimmed to display only the contained location data.  `eq_create_label()` creates an HTML snippet that be tied to the popup parameter of `leaflet` to display additional information related to the selected event

This example demonstrates both of these functions displaying all the events that have occurred in Mexico since 2000 

```{r eq_mapping_example, message = FALSE}
filename <- system.file("extdata","signif.txt", package = "Capstone")
data <- readr::read_delim(filename, delim = "\t")

eq_clean<-eq_clean_data(data)

eq_Mexico <- dplyr::filter(eq_clean,COUNTRY == "MEXICO" & lubridate::year(DATE) >2000) %>%
  dplyr::mutate(popup_text = eq_create_label(.))
eq_map(eq_Mexico, annot_col = "popup_text")
```
